---
title: "Statistical inference - GSS survey analysis for health inequity"
subtitle: "Part 1 - Categorical data"
author: "Anna Loznevaia"
date: "`r Sys.Date()`"
output: 
    html_document:
      theme: journal
      highlight: tango
      toc: true
      toc_float:
        collapsed: false
        smooth_scroll: true
      toc_depth: 6
      fig_width: 8
      fig_height: 5
      fig_caption: true
bibliography: references.bib
link-citations: true
editor_options: 
  markdown: 
    wrap: 72
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(survey)
library(srvyr)
library(questionr)
library(stringr)
library(reshape2)
library(ggpubr)
```

### Load data

I've uploaded a cross-sectional survey file from GSS web site
[\<https://gss.norc.org/get-the-data/stata/\>](https://gss.norc.org/get-the-data/stata/){.uri}
and loaded the data to DB (you can get the transfer script DB_transfer.R
from my github account). For the purpose of replicability of this
project I created three light separate data files which can be easily
downloaded (script for extracting data - DB_extract.R) - two for the
years prior 2021 (categorical and numerical inference) and one for the
year 2021.

```{r load-data}
load("gss2021.RData")
load("gss.RData")
```

------------------------------------------------------------------------

## Part 1: Data

The General Social Survey (GSS) is a nationally representative survey of
adults in the United States conducted since 1972. The GSS collects data
on contemporary American society in order to monitor and explain trends
in opinions, attitudes and behaviors. The GSS has adapted questions from
earlier surveys, thereby allowing researchers to conduct comparisons for
up to 80 years. More information can be found on the web site of the GSS
project <https://gss.norc.org/About-The-GSS>

To safeguard the health of staff and respondents during the COVID-19
pandemic, the 2021 GSS data collection used a mail-to-web methodology
instead of its traditional in-person interviews. Research and
interpretation done using the data should take extra care to ensure the
analysis reflects actual changes in public opinion and is not unduly
influenced by the change in data collection methods. For more
information on the 2021 GSS methodology and its implications, please
visit <https://gss.norc.org/Get-The-Data>

GSS collects data using multi-stage survey design, which needs to be
considered in order to generalize to the whole US population. To make my
results **generalizable** I'm going to use the survey analysis library
<https://cran.r-project.org/web/packages/survey/index.html>

This is observational study so only **correlational inference** can be
made.

------------------------------------------------------------------------

## Part 2: Research question

### Health inequity

The bases for inequalities, or determinants of health, are many.
Margaret Whitehead developed the concept of health equity adopted by
EURO/WHO.

Whitehead defines health inequities as "differences in health which are
not only unnecessary and avoidable but, in addition, are considered
unfair and unjust."

According to Whitehead, health inequalities determined by these
categories are more likely to be considered avoidable and unfair:

-   health-damaging behavior in which the degree of choice of lifestyles
    is severely restricted;

-   exposure to unhealthy, stressful living and working conditions;

-   inadequate access to essential health services and other basic
    services;

-   natural selection, or health-related social mobility, involving the
    tendency for sick people to move down the social scale.
    [@Carter-Pokras2002]

Within the US healthcare system these differences have most convincingly
been demonstrated across racial and ethnic lines (whites vs minorities);
however, disparities based on other categorizations have also been
described, including geography (urban vs rural), gender (male vs
female), socioeconomic status (poor vs nonpoor), and age (nonelderly vs
elderly).[@Gibbons2005]

**COVID-19 impact.**

Although the COVID‑19 virus infects those exposed indiscriminately,
exposure risk and the severity of its health, social and economic
impacts are not being felt equally.

COVID‑19 and its containment measures have exacerbated current and
created new vulnerabilities through three key mechanisms: unequal
socioeconomic impacts arising from both the health effects of COVID‑19
and their inequities and COVID‑19 containment measures; and the
bidirectional effects between the unequal socioeconomic impacts of
COVID‑19 and non-COVID‑19-related health inequities.
<https://www.euro.who.int/en/health-topics/health-determinants/social-determinants/publications/2020/health-inequity-and-the-effects-of-covid19-2020>

I'm going to look at the relationship between socioeconomic status
(defined by **`class`** variable) and health (defined by **`health`**
variable) using GSS cross-sectional data. Interesting question for me is
how this trend changes over time and especially in the 2021 year data.

-   **health**

    Would you say your own health, in general, is excellent, good, fair,
    or poor?

-   **class**

    If you were asked to use one of four names for your social class,
    which would you say you belong in: the lower class, the working
    class, the middle class, or the upper class?

------------------------------------------------------------------------

## Part 3: Exploratory data analysis

Creating a survey design object using instructions from codebook:

`cluster VPSU; strata VSTRAT; weight WTSSPS /* For years before 2021, use WTSSALL or WTSSNR */;`

```{r design_gss}

#for years prior 2021
gss_a_design <-svydesign(id = ~vpsu, 
                           strata = ~vstrat, 
                           data = subset(gss, 
                                         !is.na(class) & 
                                         class!="no class" & 
                                         !is.na(health)), 
                           weights = ~wtssall, 
                           nest = TRUE)

#removing empty label for class = "no class"
gss_a_design <- update(gss_a_design,class=droplevels(class))

#for 2021
gss2021_design <-svydesign(id = ~vpsu, 
                           strata = ~vstrat, 
                           data = subset(gss2021, 
                                         !is.na(class) & 
                                         class!="no class" & 
                                         !is.na(health)), 
                           weights = ~wtssnrps, 
                           nest = TRUE)

#requred for this particular design
options(survey.lonely.psu="remove")
```

I'm going to create object using `srvyr` library to use lazy tidyverse
syntax whenever I want to

```{r srvyr_obj}

gss_srvyr_design <- as_survey(gss_a_design)
gss2021_srvyr_design <- as_survey(gss2021_design)
```

Instead of ggplot I'm using `ggsurvey` from the library `questionr` for
some of the plots in this project. It should take care of weights
mapping and scaling. I compared it with manual weights scaling + option
"weight" in `ggplot` and didn't find any differences, so I will use it
as a more succinct form.

### Health distribution

Compare weighted and Unweighted health distribution for year 2021

```{r health_2021_distr_comparison, fig.height=3, fig.width=9}

gss2021 %>%
  na.omit() %>% 
  ggplot() +
  theme_light() +
  geom_bar( aes(x=health,fill = "Sample")) +
  geom_bar( aes(x=health,weight=wtssnrps,fill = "Population"), alpha=0.5) +
  scale_fill_manual(name = "Distribution", 
                     values = c(
                       "Sample" = "#0072b2", 
                       "Population" = "#e69f00"
                        )) 
```

### Health distribution by class

#### Raw contingency tables

Unweighted amount of observations for main dataset (years prior to 2021)

```{r health_counts_raw_all}

gss %>%
  filter(!is.na(health),!is.na(class),class!="no class") %>%
  group_by(class,health) %>% 
  count() %>% 
  spread(key = health, value = n)


```

There are enough observations in each group to make inference by chi
squared test

**Contingency table for 2021**

```{r health_counts_raw_2021}

gss2021 %>%
  filter(!is.na(health),!is.na(class)) %>%
  group_by(class,health) %>% 
  count() %>% 
  spread(key = health, value = n)
```

We can notice that there are not enough observations in the upper-class
& poor group to make inference based on the chi squared test.

#### Design weighted tables

We can see the difference with the original counts from the raw tables

**Contingency table for the main dataset**

```{r health_class_des}

#header for contingency table
gss_s <- subset(gss, select=c(class,health), !is.na(class) & class!="no class")
gss_s <- droplevels(gss_s)
rows <- sort(unique(gss_s$class))
cols <- sort(unique(gss_s$health))

#creating new tables with counts
t_a <-svytotal(~interaction(health,class),design=gss_a_design)
ftable(t_a, rownames=list(health=cols,class=rows))
```

**Contingency table for 2021**

```{r health_class_des_2021}

#creating new tables with counts
t_2021 <-svytotal(~interaction(health,class),design=gss2021_design)
ftable(t_2021, rownames=list(health=cols,class=rows))
```

It's even worse after accounting for weights, so we can not use the chi
squared test for the year 2021 separately.

### Health distribution by class and years

```{r health_dist_class_year_all}

#calculating percentages
gss_srvyr_design <- gss_srvyr_design %>% 
  group_by(year, class, health) %>% 
  mutate(year_health_count = n()) %>% 
  group_by(year,class) %>% 
  mutate(proc_health_year = year_health_count/n())

gss_srvyr_design %>% 
  ggsurvey() + 
  aes(x = year,y=proc_health_year,color=health) + 
  geom_smooth(size=1) +
  facet_wrap(~class) + 
  theme_light() + 
  scale_color_brewer(palette = "Set1") +
  labs(y="Health proportion", x="")
```

**Health distribution by class 2021**

```{r health_dist_class_2021, fig.height=3, fig.width=9}

gss_srvyr_design %>% 
  ggsurvey() + 
  aes(x = class,fill=health) + 
  geom_bar(position = "fill") +
  labs(y = "Health proportion") +
  coord_flip() + 
  theme_light() +
  scale_fill_brewer(palette = "Set2")
```

------------------------------------------------------------------------

## Part 4: Inference

### Chi squared

Condition for inference using chi squared is satisfied only for the main
dataset:

-   independence - observations are independent between and within the
    groups, data is not paired (from GSS survey design), sample size
    \<10% of the population

-   number of observations in each of the groups \>5.

H~0~: distribution of health is independent of class, all variations is
due to chance $\pi_{1} = \pi_{2}= \pi_{3}= \pi_{4}$

H~1~: distribution of health is not independent of class
$\pi_{1} \neq \pi_{2} \neq \pi_{3} \neq \pi_{4}$

```{r health_class_chi}

chi<-svychisq(~class + health, design = gss_a_design, statistic = 'Chisq')
chi 
```

P-value is almost zero which allow us to reject H~0~ and conclude that
distribution of health is not independent of the class

### Mean + CI

Estimated proportions and their 95% CI

```{r health_class_mean}

mean_tab <- svyby( formula = ~health,
                   by = ~class,
                   design = gss_a_design,
                   FUN = svymean,
                   na.rm = TRUE )

#plot friendly dataset - from wide to long 

colnames(mean_tab) <- c("class","excellent","good","fair","poor","se.excellent","se.good","se.fair","se.poor")
dm1 <- melt(mean_tab[,c("class","excellent","good","fair","poor")], id=c("class"))
dm2 <- melt(mean_tab[,c("class","se.excellent","se.good","se.fair","se.poor")], id=c("class"))
colnames(dm1) <- c("class", "health", "p")
dm2$variable <- str_replace(dm2$variable, "se.", "")
colnames(dm2) <- c("class", "health", "se")
mean_tab <- merge(dm1, dm2)

mean_tab <- mean_tab %>%
  mutate(p_l = p - 1.96*se, p_h = p + 1.96*se )
mean_tab
```

```{r health_class_mean_CI}
# Construct a bar plot of average income by sex
mean_tab %>% 
  ggplot() +
    aes(x = health, y = p, ymin = p_l, ymax = p_h, fill = health) +
    geom_col() +
    geom_errorbar(width = 0.7) + 
    theme_light() + 
    scale_fill_brewer(palette = "Set2")+
    labs(y="Mean health") +
    facet_wrap(~class) +
    theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```

### Statistic yearly dynamic

This data isn't very reliable because some groups has less then 5
observations but it is anyway interesting to look at. For each year I
calculated 4 different types of statistic:

The default (`statistic="F"`) is the Rao-Scott second-order correction.
The p-values are computed with a Satterthwaite approximation to the
distribution. The alternative `statistic="Chisq"` adjusts the Pearson
chisquared statistic by a design effect estimate and then compares it to
the chisquared distribution it would have under simple random sampling.
The `statistic="Wald"` test is that proposed by Koch et al (1975) and
used by the SUDAAN software package. It is a Wald test based on the
differences between the observed cell counts and those expected under
independence. The adjustment given by `statistic="adjWald"` reduces the
statistic when the number of PSUs is small compared to the number of
degrees of freedom of the test. Rao and Thomas (1990) compare these
tests and find the adjustment beneficial. From
<https://r-survey.r-forge.r-project.org/survey/html/svychisq.html>

```{r class_year_dyn }
years <- unique(gss$year)
f_all <- NULL

for (i in years) {
  gss_design <-svydesign(id = ~vpsu, 
                           strata = ~vstrat, 
                           data = subset(gss, 
                                          !is.na(class) & 
                                          class!="no class" & 
                                          year == i &
                                          !is.na(health)), 
                           weights = ~wtssall, 
                           nest = TRUE)

    gss_design <- update(gss_design,class= droplevels(class))
    chi1<-svychisq(~class + health, design = gss_design, statistic = 'Chisq')
    chi2<-svychisq(~class + health, design = gss_design, statistic = 'F')
    chi3<-svychisq(~class + health, design = gss_design, statistic = 'Wald')
    chi4<-svychisq(~class + health, design = gss_design, statistic = 'adjWald')
    
    df<-data.frame(year=i,
                   Chisq_st = chi1$statistic,
                   F_st = chi2$statistic,
                   Wald_st = chi3$statistic,
                   adjWald_st = chi4$statistic )
    row.names(df) <- i
    f_all <- rbind(f_all,df)
}

#adding 2021
   chi1<-svychisq(~class + health, design = gss2021_design, statistic = 'Chisq')
   chi2<-svychisq(~class + health, design = gss2021_design, statistic = 'F')
   chi3<-svychisq(~class + health, design = gss2021_design, statistic = 'Wald')
   chi4<-svychisq(~class + health, design = gss2021_design, statistic = 'adjWald')
    
    df<-data.frame(year=2021,
                   Chisq_st = chi1$statistic,
                   F_st = chi2$statistic,
                   Wald_st = chi3$statistic,
                   adjWald_st = chi4$statistic )
    row.names(df) <- i
    f_all <- rbind(f_all,df)

f_all %>% 
  ggplot() +
    geom_line(aes(x=year,y=Chisq_st,color="Chisquared"),size=1) + 
    geom_line(aes(x=year,y=2*F_st,color="2X Rao-Scott"),size=1) +
    geom_line(aes(x=year,y=2*Wald_st,color="2X Wald"),size=1) +
    geom_line(aes(x=year,y=2*adjWald_st,color = "2X adjusted Wald"),size=1)+
    scale_color_manual(name = "statistics", 
                     values = c(
                       "Chisquared" = "#0072b2", 
                       "2X Rao-Scott" = "#e69f00",
                       "2X Wald" = "#56b4e9",
                       "2X adjusted Wald" = "#cc79a7"
                       )) +
    geom_point(aes(x=year,y=Chisq_st)) +
    geom_text(
      aes(x=year,y=Chisq_st),
      label=f_all$year, 
      position = position_nudge(x = 1,y = -1),  
      size=3
    ) 
```

------------------------------------------------------------------------

## Part 5: Opinion

GSS provides a rich source of information about opinion trends. As a
conclusion I would like to look at how the situation with health
disparity is reflected in mass consciousness by one example:

**natfare**

> We are faced with many problems in this country, none of which can be
> solved easily or inexpensively. I'm going to name some of these
> problems, and for each one I'd like you to name some of these
> problems, and for each one I'd like you to tell me whether you think
> we're spending too much money on it, too little money, or about the
> right amount. First (READ ITEM A) . . . are we spending too much, too
> little, or about the right amount on (ITEM)? K. Welfare

#### Distribution of the opinions about welfare spending

```{r welfare_op }

#plot 1 - all years prior 2020
p1 <- gss_srvyr_design %>% 
  na.omit() %>% 
  ggsurvey() + 
    aes(x = class, fill = natfare) + 
    geom_bar(position = "fill") +
    theme_light() + 
    scale_color_brewer(palette = "Set1")+
    theme(axis.title.x=element_blank(),legend.title = element_blank())

#plot 2 - 2021 data
p2 <- gss2021_srvyr_design %>% 
  na.omit() %>% 
  ggsurvey() + 
    aes(x = class, fill = natfare) + 
    geom_bar(position = "fill") +
    theme_light() + 
    scale_color_brewer(palette = "Set1")+
    theme(axis.title.x=element_blank(),legend.title = element_blank()) 

#arranging together
ggarrange(
  p1 + rremove("x.text") + rremove("grid"), 
  p2 + rremove("grid"),
  labels = c("Before 2020", "2021 dataset"), hjust=-0.5,
  common.legend = TRUE, legend = "right", 
  ncol = 1, nrow = 2 )

```

#### Trend by years

```{r welfare_op_dyn, fig.height=4}

#calculating percentages for the graph

out_fare <- svyby(formula = ~natfare, 
                    by = ~year, 
                    design = gss_a_design, 
                    FUN = svymean,
                    na.rm = TRUE
)

#wide to long format
colnames(out_fare) <- c("year","Too little","About right","Too much")
out_fare <- melt(out_fare[,c("year","Too little","About right","Too much")], id=c("year"))
colnames(out_fare) <- c("year","natfare","p")

out_fare2021 <- svymean(x = ~natfare, design = gss2021_design, na.rm = TRUE )
out_fare2021 <- as.data.frame(out_fare2021)
out_fare2021$year <- 2021
out_fare2021$natfare <- c("Too little","About right","Too much")
out_fare2021 <- subset(out_fare2021, select = -c(SE) )
colnames(out_fare2021) <- c("p","year","natfare")

#combining all years
out_fare_tot <- rbind(out_fare, out_fare2021)

out_fare_tot %>% 
  ggplot() +
    aes(x=year, y=p, color=natfare) +
    geom_line() + 
    geom_point() +
    theme_light()+
    scale_color_brewer(palette = "Set1") +
    labs(y="proportion",color="Welfare spending") +
    theme(axis.title.x=element_blank())
```

I think it is getting better, but still while our planet facing
ecological meltdown and COVID-19 pandemic is far from over - we will
have to deal with the dire socioeconomic consequences in many years to
come, a lot of non-essential (softly speaking) projects such as space
tourism attract a lot of resources and public attention.
