---
title: "Statistical inference - GSS survey analysis: Part 2 - Numerical data"
subtitle: "Gender pay gap"
author: "Anna Loznevaia"
date: "`r Sys.Date()`"
output: 
    html_document:
      theme: journal
      highlight: tango
      toc: true
      toc_float:
        collapsed: false
        smooth_scroll: true
      toc_depth: 6
      fig_width: 9
      fig_height: 7
      fig_caption: true
bibliography: references.bib
link-citations: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r}

knitr::opts_chunk$set( warning=FALSE, message=FALSE)
```

## Setup

### Load packages

```{r load-packages, warnings = FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(survey)
library(srvyr)
library(questionr)
library(scales)
```

Part 1 - Inference on categorical data - health inequity 2021
<https://github.com/Myoth13/GSS/blob/main/GSS_health_inequity_2021.html>

### Load data

I've uploaded a cross-sectional survey file from GSS web site
[<https://gss.norc.org/get-the-data/stata/>](https://gss.norc.org/get-the-data/stata/){.uri}
and loaded the data to a DB (transfer script
<https://github.com/Myoth13/GSS/blob/main/DB_transfer.R>). For the
purpose of replicability of this project I created three light separate
data files which can be easily downloaded (script for extracting data -
<https://github.com/Myoth13/GSS/blob/main/DB_extract.R>) - two for the
years prior 2020 (categorical and numerical inference) and one for the
year 2021.

```{r load-data}
load("gss_num.RData")
```

------------------------------------------------------------------------

## Part 1: Data

The General Social Survey (GSS) is a nationally representative survey of
adults in the United States conducted since 1972. The GSS collects data
on contemporary American society in order to monitor and explain trends
in opinions, attitudes and behaviors. The GSS has adapted questions from
earlier surveys, thereby allowing researchers to conduct comparisons for
up to 80 years. More information can be found on the web site of the GSS
project <https://gss.norc.org/About-The-GSS>

GSS collects data using multi-stage survey design, which needs to be
considered in order to generalize to the whole US population. To make my
results **generalizable** I'm going to use the survey analysis library
<https://cran.r-project.org/web/packages/survey/index.html>

This is observational study so only **correlational inference** can be
made.

This particular dataset is not very good for analyzing numerical data -
the study design intended to be categorical based (opinions, attitudes
and behaviours trends). So even supposed to be numerical data like
respondent's income is originally collected as categorical data and then
post processed which imply inevitable loss of information. And second -
the dataset is not large enough for in-depth (based on many
sub-variables) analysis and inference.

------------------------------------------------------------------------

## Part 2: Research question

#### Gender pay gap

According to Bureau of Labor Statistics data, in 2020, women's annual
earnings were 82.3% of men's, and the gap is even wider for many women
of color. Though women only made 57 cents per dollar earned by men in
1973 when this [Department of Labor
PSA](https://www.youtube.com/watch?v=n00xZ_mKQgk%5d) was made, progress
has stalled and we're still far from closing the pay gap.

[Compared with white men with the same
education](https://www.bls.gov/cps/demographics.htm#education), Black
and Latina women with only a bachelor's degree have the largest gap at
65%, and Black women with advanced degrees earn 70% of what white men
with advanced degrees earn. Educational attainment is not enough to
close gender earnings gaps. In fact, most women with advanced degrees
earn less than white men, on average, with only a bachelor's degree.

Unfortunately, the pandemic stalled gains made toward closing the pay
gap, and layoffs and a lack of child care have forced many women out of
the workforce entirely. In February 2021, women's labor force
participation rate was 55.8% -- the same rate as April 1987. And women
of color and those working in low-wage occupations [have been the most
impacted](https://www.youtube.com/watch?v=xyX1cjxCG7g).

From
<https://blog.dol.gov/2021/03/19/5-facts-about-the-state-of-the-gender-pay-gap>

I would like to see the general yearly dynamic in the gender pay gap as
well as the influence of some other factors such as class, race and age
on this dynamic. In my analysis I tried to select variables with the
maximum available information across the dataset.

-   **conrinc**

> Inflation-adjusted personal income.

> Notes: Income variables (INCOME, RINCOM77, RINCOM82, RINCOM86,
> RINCOM91) are recoded in six-digit numbers and converted to 2000
> dollars. The collapsed numbers above are for convenience of display
> only. Since this variable is based on categorical data, **income is
> not continuous**, but based on categorical mid-points and imputations.
> For details see GSS Methodological Report No. 101.

The data is not available for 2021.

-   **sex**

> The variable SEX in face-to-face interviews, interviewers
> traditionally coded SEX based on interviewer observation

-   **degree**

> Respondent's degree

-   **class**

> If you were asked to use one of four names for your social class,
> which would you say you belong in: the lower class, the working class,
> the middle class, or the upper class?

-   **race**

> What race do you consider yourself?

-   **age_group**

arbitrary derivative from the variable `age`

> Respondent's age

------------------------------------------------------------------------

## Part 3: Exploratory data analysis

**Missing data**: 2010 data has no real information on `conrinc`
variable, so I've decided not to use it in the analysis. Year 2021 has
no `conrinc` variable.

```{r conrinc_2010}
gss_num %>% 
  filter(year==2010,!is.na(conrinc)) %>% 
  select(conrinc) %>% 
  summary()

```

**Survey design**

```{r desing_num}

#scaling weights to US population size
gss_num$wtssall_scaled <- gss_num$wtssall * 258000000 / nrow( gss_num )

gss_num_design <-svydesign(id = ~vpsu, 
                           strata = ~vstrat, 
                           data = subset(gss_num, !is.na(conrinc) & year!=2010
                                         & class!="no class"), 
                           weights = ~wtssall, 
                           nest = TRUE)

#removing empty label for class = "no class"
gss_num_design <- update(gss_num_design,class=droplevels(class))

#requred for this particular design
options(survey.lonely.psu="remove")
```

I'm going to create object using `srvyr` library to use lazy tidyverse
syntax whenever I want to

```{r srvyr_obj}

gss_num_srvyr_design <- as_survey(gss_num_design)
```

### Income distribution

Compare weighted and unweighted distribution

```{r income_distr, fig.height=5}
# Create a weighted histogram with ggplot
gss_num %>%
  na.omit() %>% 
  ggplot() +
    theme_light() +
    geom_histogram( aes(x=conrinc,color = "Sample",fill = "Sample"), 
                    binwidth = 7000, alpha=0.5) +
    geom_histogram( aes(x=conrinc,weight=wtssall,color = "Population",fill = "Population"), 
                    binwidth = 7000) +
    labs(x = "Income") +
    scale_color_manual(name = "Distribution", 
                      values = c(
                        "Sample" = "gold", 
                        "Population" = "navy"
                      )) +
    scale_fill_manual(name = "Distribution", 
                      values = c(
                        "Sample" = "gold", 
                        "Population" = "transparent"
                      )) +
  scale_x_continuous(labels = comma)
```

There is a noticeable difference between weighted and unweighted
distribution. Without accounting for this difference results will be
biased and variance calculated incorrectly, which can result in the
wrong inference.

**Summary statistics**

```{r conrinc_summary}
summary(gss_num$conrinc)
```

Distribution is very skewed. I would prefer to use median whenever
possible.

### **I**ncome distribution by gender

```{r income_distr_s, fig.height=5}

gss_num_srvyr_design %>% 
  ggsurvey() +
    theme_light() +
    geom_histogram( aes(x=conrinc,color = sex,fill = sex), binwidth = 7000, alpha=0.5) +
    labs(x = "Income") +
    scale_x_continuous(labels = comma) +
    scale_color_brewer(palette = "Set1")+
    scale_fill_brewer(palette = "Set1") +
    facet_wrap(~sex) 

```

### Survey-weighted medians

#### Quantiles with 95% CI

```{r conrinc_quantiles_des}
svyquantile(x = ~conrinc, 
            design = gss_num_design, 
            na.rm = TRUE, 
            quantiles = c(0.01, 0.25, 0.5, 0.75,0.99))
```

For `survey.design` objects the default is `interval.type="Wald"`. A 95%
Wald confidence interval is constructed for the proportion below the
estimated quantile. The inverse of the estimated CDF is used to map this
to a confidence interval for the quantile.

#### By gender

-   Survey-weighted median income by `sex`

```{r conrinc_s_median_table}
# Compute the survey-weighted medians by sex
svyby(formula = ~conrinc, 
          by = ~sex, 
          design = gss_num_design, 
          FUN = svyquantile, 
          na.rm = TRUE, 
          quantiles = 0.5, 
      ) 
```

Median income between genders differs. Variance is homogeneous across
the groups.

-   Survey-weighted median income by `sex` and `year` with 95% CI

```{r conrinc_s_y_median}
out_s <- svyby(formula = ~conrinc, 
                    by = ~year+sex, 
                    design = gss_num_design, 
                    FUN = svyquantile,
                    quantiles=0.5,
                    na.rm = TRUE,
                    ci=TRUE, vartype="ci"
)

# Construct a bar plot of average income by gender
out_s %>% 
  ggplot() +
    aes(x = sex, y = conrinc, ymin = ci_l, ymax = ci_u, fill = sex) +
    geom_col() +
    geom_errorbar(width = 0.7) + 
    theme_light() + 
    scale_fill_brewer(palette = "Set2")+
    labs(y="Median income") +
    facet_wrap(~year) +
    theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

```

On this bar plot we can see that the median difference is getting
smaller with years. To better see the dynamics I made a linear plot

```{r conrinc_s_y_dyn, fig.height=4}
out_s %>% 
  ggplot() +
  geom_line(aes(x=year, y=as.numeric(conrinc), color=sex),size=1) + 
  theme_light()+
  scale_color_brewer(palette = "Set1") +
  labs(y="median income") 

```

#### By gender and degree

If we look at the counts of cases in each group (`sex`+ `year`+`degree`)
we can see that there is not enough data to make inference. But anyway
it would be interesting to look at the graph.

```{r conrinc_s_y_d_count}
out_s_d_tot <- svytable(~sex+degree+year, design = gss_num_design)
head(as.data.frame(out_s_d_tot))
```

Graph

```{r conrinc_s_y_d_dyn, fig.height=6}

out_s_d <- svyby(formula = ~conrinc, 
                    by = ~year+sex+degree, 
                    design = gss_num_design, 
                    FUN = svyquantile,
                    quantiles=0.5,
                    na.rm = TRUE
)

out_s_d %>% 
  ggplot() +
    geom_smooth(aes(x=year, y=as.numeric(conrinc), color=sex),size=1) + 
    theme_light()+
    scale_color_brewer(palette = "Set1") +
    labs(y="median income") +
    facet_wrap(~degree)
```

We can see that dynamics is not homogeneous across the groups and
variance in the graduate male group is too high.

#### By gender and class

If we look at the counts of cases in each group (`sex`+ `year`+`class`)
we can see that there is not enough data to make inference.

```{r conrinc_s_y_cl_count}
out_s_cl_tot <- svytable(~sex+class+year, design = gss_num_design)
head(as.data.frame(out_s_cl_tot))
```

Graph

```{r conrinc_s_y_cl_dyn, fig.height=6}

out_s_cl <- svyby(formula = ~conrinc, 
                    by = ~year+sex+class, 
                    design = gss_num_design, 
                    FUN = svyquantile,
                    quantiles=0.5,
                    na.rm = TRUE
)

out_s_cl %>% 
  ggplot() +
    geom_smooth(aes(x=year, y=as.numeric(conrinc), color=sex),size=1) + 
    theme_light()+
    scale_color_brewer(palette = "Set1") +
    labs(y="median income") +
    facet_wrap(~class)
```

Yearly dynamics is very different for different socioeconomic classes -
almost the same level for the low and working class, some difference in
the middle class and too much variance in the upper class to make any
conclusions.

#### By gender and race

If we look at the counts of cases in each group (`sex`+ `year`+`race`)
we can see that there is not enough data to make inference.

```{r conrinc_s_y_r_count}
out_s_r_tot <- svytable(~sex+race+year, design = gss_num_design)
head(as.data.frame(out_s_r_tot))
```

Graph

```{r conrinc_s_y_r_dyn, fig.height=4}

out_s_r <- svyby(formula = ~conrinc, 
                    by = ~year+sex+race, 
                    design = gss_num_design, 
                    FUN = svyquantile,
                    quantiles=0.5,
                    na.rm = TRUE
)

out_s_r %>% 
  ggplot() +
    geom_smooth(aes(x=year, y=as.numeric(conrinc), color=sex),size=1) + 
    theme_light()+
    scale_color_brewer(palette = "Set1") +
    labs(y="median income") +
    facet_wrap(~race)
```

Way too much variance in the other race group (due to the lack of
information in the dataset), but looking at the other two seems like the
only privileged group now is white males.

#### By gender and age

If we look at the counts of cases in each group (`sex`+
`year`+`age_group`) we can see that there is not enough data to make
inference.

```{r conrinc_s_y_age_count}
out_s_age_tot <- svytable(~sex+age_group+year, design = gss_num_design)
head(as.data.frame(out_s_age_tot))
```

Graph

```{r conrinc_s_y_age_dyn, fig.height=6}

out_s_age <- svyby(formula = ~conrinc, 
                    by = ~year+sex+age_group, 
                    design = gss_num_design, 
                    FUN = svyquantile,
                    quantiles=0.5,
                    na.rm = TRUE
)

out_s_age %>% 
  ggplot() +
    geom_smooth(aes(x=year, y=as.numeric(conrinc), color=sex),size=1) + 
    theme_light()+
    scale_color_brewer(palette = "Set1") +
    labs(y="median income") +
    facet_wrap(~age_group)
```

Looks like since the 2000th in the middle aged groups things are not
getting much better.

## Part 4: Inference

T-test and ANOVA inference is mean-based. Conditions for both are
satisfied:

-   independence - observations are independent between and within the
    groups, data is not paired (from GSS survey design), sample size
    \<10% of the population

-   distribution of the data within each group is not normal, but sample
    size is big enough.

-   variability across the groups is about equal

### T test

H~0~: there is no difference between mean income of males and females
$\mu_{male} = \mu_{female}$

H~1~: there is a difference between mean income of males and females
$\mu_{male} \neq \mu_{female}$

```{r ttest}
##Inference for quantitative data ###Survey-weighted t-test
svyttest(formula = conrinc~sex,
         design = gss_num_design)


```

### ANOVA

For each of the variable we have a pair of hypotheses

H~0~: there is no difference between mean income
$\mu_{1} = \mu_{2} = \mu_{...} =\mu_{i}$

H~1~: there is a difference between mean income
$\mu_{1} \neq \mu_{2} \neq \mu_{...} \neq \mu_{i}$

To extract ANOVA table from survey design based model I have to use GLM
model first.

```{r conrinc_degree_anov_surv}
# Extracting ANOVA table for income by degree and gender
gss.glm1 <- svyglm(conrinc~sex+degree+race+class+age_group, design=gss_num_design)
atable <- anova(gss.glm1)
atable
```

Comparing to generic anova with `weights` parameter

```{r conrinc_degree_anov_generic}
atable2 <- aov(conrinc~degree+sex+race+class+age_group, data = gss_num, weights = wtssall)
summary(atable2)
```

Both variants shows that all variables are significant with p-value
approaching zero. Thus we can reject both H~0~ hypotheses.
