---
title: "Survey - categorical data"
output: html_notebook
---

Survey design - (cluster VPSU; strata VSTRAT; weight WTSSPS /\* For years before 2021, use WTSSALL or WTSSNR \*/;)

```{r}
library(dplyr)
library(ggplot2)
library(survey)
library(questionr)

load("gss.RData")

gss_design <-svydesign(id = ~vpsu, 
                           strata = ~vstrat, 
                           data = subset(gss,
                                         class!="no class" &
                                        !is.na(class)), 
                           weights = ~wtssall, 
                           nest = TRUE)

gss_design <- update(gss_design,class=droplevels(class)) 

head(gss_design)
```

Estimated population size. In this case weights are scaled to sample size

```{r}
gss %>% 
  summarize(N_hat = sum(wtssall))
```

Number of strata

```{r}
gss %>%
  summarize(n_starta = n_distinct(vstrat))
```

Number of clusters

```{r}
gss %>%
  summarize(n_clusters = n_distinct(vstrat,vpsu))
```

Sample sizes in clusters

```{r}
gss %>%
  count(vstrat,vpsu)
```

Construct a frequency table of Classes

```{r}
tab_cl <- svytable(~class, design = gss_design)
tab_cl
```

Add proportions to table

```{r}
tab_cl <- tab_cl %>%
  as.data.frame() %>%
  mutate(Prop = Freq/sum(Freq))
tab_cl
```

Create a barplot

```{r}
tab_cl %>% ggplot() + 
  aes(x = class, y = Prop) +
  theme(axis.text.x = element_text(size=11, angle = 90, vjust = 0.5, hjust=1)) +
  geom_col()

```

Same plot using ggsurvey

```{r}
ggsurvey(gss_design) + 
  aes(x = class) + 
  theme(axis.text.x = element_text(size=11, angle = 90, vjust = 0.5, hjust=1)) +
  geom_bar() 
```

Construct a frequency table of Classes + Health

```{r}
tab_cl_h <- svytable(~class+health, design = gss_design)
tab_cl_h
```

Add conditional proportions to table

```{r}
tab_cl_h <- tab_cl_h %>%
  as.data.frame() %>% 
  group_by(class) %>% 
  mutate(sum_class=sum(Freq),prop_health = Freq/sum_class) %>% 
  ungroup()
tab_cl_h
```

Create a segmented bar graph of the conditional proportions in tab_DH_cond

```{r}
tab_cl_h %>%  
  ggplot() +
    aes(x = class, y = prop_health, fill = health) + 
    geom_col() + 
    coord_flip()

```

ggsurvey variant

```{r}
gss_design %>%  
  na.omit() %>% 
  ggsurvey() + 
    aes(x = class, fill = health) + 
    geom_bar(position = "fill") +
    coord_flip()
```

Summary for class + health

```{r}
options(survey.lonely.psu="remove")
svytotal(x = ~interaction(class, health),
                       design = gss_design,
                       na.rm = TRUE)

```

Estimate the means for class + health

```{r}
svymean(x = ~interaction(class, health),
                     design = gss_design,
                     na.rm = TRUE)
```
