---
title: "Survey - numerical data"
output: html_notebook
---

Survey design - (cluster VPSU; strata VSTRAT; weight WTSSPS /\* For years before 2021, use WTSSALL or WTSSNR \*/;)

```{r message = FALSE}
library(dplyr)
library(ggplot2)
library(survey)
library(questionr)

load("gss_num.RData")

gss_num_design <-svydesign(id = ~vpsu, 
                           strata = ~vstrat, 
                           data = subset(gss_num, !is.na(conrinc) & year!=2010), 
                           weights = ~wtssall, 
                           nest = TRUE)

#requred for this particular design
options(survey.lonely.psu="remove")

head(gss_num_design)
```

Estimated population size. In this case weights are scaled to sample size

```{r}
gss_num %>% 
  summarize(N_hat = sum(wtssall))
```

Number of strata

```{r}
gss_num %>%
  summarize(n_strata = n_distinct(vstrat))
```

Number of clusters

```{r}
gss_num %>%
  summarize(n_clusters = n_distinct(vstrat,vpsu))
```

Sample sizes in clusters

```{r}
gss_num %>%
  count(vstrat,vpsu)
```

Distribution of income by sex (weights scaling)

```{r}
gss_num %>%
  filter(!is.na(conrinc)) %>%
  group_by(sex) %>%
  mutate(wtssall_std = wtssall/sum(wtssall)) %>%
  ggplot() + 
    aes(x = conrinc, weight = wtssall_std) + 
    geom_histogram(binwidth = 1000,  fill = "magenta") +
    labs(x = "Income") + 
    facet_wrap(~sex, labeller = "label_both", scales = "free")
```

Distribution of income by sex (using questionr)

```{r}
gss_num_design %>%  
  ggsurvey() + 
    aes(x = conrinc) + 
    geom_density(bw = 0.6,  color = "gold") +
    labs(x = "Income") + 
    facet_wrap(~sex, labeller = "label_both", scales = "free")
```

Boxplot of the distribution

```{r}
gss_num_design %>%  
  ggsurvey() + 
  aes(x = conrinc, y = sex) + 
  geom_boxplot() 
```

Compute the survey-weighted mean

```{r}
svymean(x = ~conrinc, 
        design = gss_num_design,
        na.rm = TRUE)
```

Compute the survey-weighted mean by sex

```{r}
svyby(formula = ~conrinc, 
      by = ~sex, 
      design = gss_num_design, 
      FUN = svymean, 
      na.rm = TRUE)
```

Compute the survey-weighted quantiles

```{r}
svyquantile(x = ~conrinc, 
            design = gss_num_design, 
            na.rm = TRUE, 
            quantiles = c(0.01, 0.25, 0.5, 0.75,0.99))
```

Compute the survey-weighted medians by sex

```{r}
svyby(formula = ~sex+conrinc, 
      by = ~sex, 
      design = gss_num_design, 
      FUN = svyquantile, 
      na.rm = TRUE, 
      quantiles = 0.5
) 
```

Construct a bar plot of average income by sex

```{r}
# Compute the survey-weighted mean by degree
out_mean <- svyby(formula = ~conrinc, 
                    by = ~sex+year, 
                    design = gss_num_design, 
                    FUN = svymean,
                    na.rm = TRUE)

# Add lower and upper columns to out
out_mean <- out_mean %>%  
  mutate(lower = conrinc - 1.96*se, 
         upper = conrinc + 1.96*se)

# Construct a bar plot of average income by degree with error bars
out_mean %>% 
  ggplot()+ 
    aes(x = sex, y = conrinc, ymin = lower, ymax = upper) +
    geom_col(fill = "gold") +
    labs(y = "Average income") +
    geom_errorbar(width = 0.7) + 
    facet_wrap(~year)
```
